---
title: "Lab: Creating Good Graphics"
author: "Tyler Davis"
format: html
number-sections: true
number-depth: 2
messages: false
errors: false
warnings: false
editor: 
  markdown: 
    wrap: sentence
---

::: callout
You can see the purpose of this assignment as well as the skills and knowledge you should be using and acquiring, in the [Transparency in Learning and Teaching (TILT)](tilt.qmd) document in this repository.
The TILT document also contains a checklist for self-reflection that will provide some guidance on how the assignment will be graded.
:::

# Graphical Critiques

The following code was [generated by Claude AI vers. Sonnet 4.5](4-types-graphics-claude.md) (with some modifications).

```{r}
#| label: fig-comparing-graphics-figures
#| fig-cap: "Four different representations of COVID vaccination rates."
#| fig-subcap: ["Caption 1", "Caption 2", "Caption 3", "Caption 4"]
#| fig-alt: ["This figure is a pie chart that displays COVID vaccination rates by age group. The age groups are colored and the percent of overall COVID vaccination rates are displayed.", "This figure is a line chart which diplays COVID vaccination rates are diplayed over a 12 week period. The x axis is mapping weeks while the y axis is mapping vaccination rate. Finally, the color is being mapped to age group.", "This figure is a stacked chart, descrbing vaccination rates over time and sorted by age group, which is the color mapping. The x axis is displaying time while the y axis is showing percent.", "This figure is a bar chart which diplays COVID vaccination rates are diplayed over a 12 week period. The x axis is mapping weeks while the y axis is mapping vaccination rate. Finally, the color is being mapped to age group."]
#| layout-ncol: 2
#| code-fold: true


# Load required libraries
suppressPackageStartupMessages({
  library(ggplot2)
  theme_set(theme_bw())
  library(dplyr)
  library(tidyr)
})

# Create sample dataset
set.seed(123)
weeks <- 1:12
age_groups <- c("18-29", "30-49", "50-64", "65+")

# Generate vaccination rate data with realistic patterns
vaccination_data <- data.frame(
  week = rep(weeks, each = 4),
  age_group = rep(age_groups, times = 12),
  vaccination_rate = c(
    # Week-by-week data for each age group
    12, 18, 25, 35,  # Week 1
    15, 22, 30, 41,  # Week 2
    18, 26, 35, 47,  # Week 3
    22, 31, 41, 54,  # Week 4
    26, 36, 47, 60,  # Week 5
    30, 41, 53, 66,  # Week 6
    34, 46, 58, 71,  # Week 7
    38, 50, 63, 75,  # Week 8
    42, 54, 67, 79,  # Week 9
    45, 58, 71, 82,  # Week 10
    48, 61, 74, 85,  # Week 11
    51, 64, 77, 87   # Week 12
  )
)

# Set factor levels for consistent ordering
vaccination_data$age_group <- factor(vaccination_data$age_group, 
                                     levels = c("18-29", "30-49", "50-64", "65+"))

# Define color palette
age_colors <- c("18-29" = "#FF6B6B", 
                "30-49" = "#4ECDC4", 
                "50-64" = "#FFD93D", 
                "65+" = "#95E1D3")

# CHART A: 3D Pie Chart (showing total vaccinations by age group)

# Calculate total vaccinations per age group (sum across weeks)
total_by_age <- vaccination_data %>%
  group_by(age_group) %>%
  summarise(total = sum(vaccination_rate))

# Create pie chart with 3D-like appearance using custom positioning
chart_a <- ggplot(total_by_age, aes(x = "", y = total, fill = age_group)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar("y", start = 0) +
  scale_fill_manual(values = age_colors) +
  labs(title = "Chart A: 3D Pie Chart - Total Vaccinations by Age Group",
       subtitle = "(Summed across all 12 weeks)",
       fill = "Age Group") +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 10, face = "italic"),
    legend.position = "right",
    legend.title = element_text(size = 11, face = "bold"),
    legend.text = element_text(size = 10)
  ) +
  # Add percentage labels
  geom_text(aes(label = paste0(round(total/sum(total)*100, 1), "%")),
            position = position_stack(vjust = 0.5),
            size = 4, fontface = "bold")

# CHART B: Line Graph - Vaccination Rates Over Time

chart_b <- ggplot(vaccination_data, aes(x = week, y = vaccination_rate, 
                                        color = age_group, group = age_group)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_color_manual(values = age_colors) +
  scale_x_continuous(breaks = 1:12) +
  labs(title = "Chart B: Line Graph - Vaccination Rates Over Time",
       x = "Week",
       y = "Vaccination Rate (%)",
       color = "Age Group")

# CHART C: Stacked Area Chart - Vaccination Rates Over Time

chart_c <- ggplot(vaccination_data, aes(x = week, y = vaccination_rate, 
                                        fill = age_group)) +
  geom_area(alpha = 0.8, position = "stack") +
  scale_fill_manual(values = age_colors) +
  scale_x_continuous(breaks = 1:12) +
  labs(title = "Chart C: Stacked Area Chart - Vaccination Rates Over Time",
       x = "Week",
       y = "Cumulative Vaccination Rate (%)",
       fill = "Age Group")


# CHART D: Grouped Bar Chart - Vaccination Rates by Week and Age Group

chart_d <- ggplot(vaccination_data, aes(x = factor(week), y = vaccination_rate, 
                                        fill = age_group)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.8) +
  scale_fill_manual(values = age_colors) +
  labs(title = "Chart D: Grouped Bar Chart - Vaccination Rates by Week and Age Group",
       x = "Week",
       y = "Vaccination Rate (%)",
       fill = "Age Group")

chart_a
chart_b
chart_c
chart_d
```

1.  Edit the chunk options above to produce helpful and descriptive captions and alt-text for each chart. Every chart you include in a document should be referenced in the text (preferably using a cross-reference) and should have a helpful caption and include at least short alt-text descriptions using the `fig-alt` quarto option.\
    Resources:

-   [Mastering Alt Text for Images and Charts](https://a11ykat.medium.com/see-the-unseen-mastering-alt-text-for-images-and-charts-6d3dbe36553b)
-   [Image Accessibility Generator](https://asuo-ai-labs.streamlit.app/Image_Accessibility) - This is a great tool but I often have to customize the output. I usually use the long text for image descriptions, which is not *best* practice but tends to work better than many of the other options available within the quarto ecosystem.

2.  Using the grammar of graphics, identify the geom, mappings, stat, positions, coordinate systems, and transformations for each graphic. Fill in the table below accordingly, leaving any cells which do not apply to the chart blank. If a mapping is a function of a statistic, fill in the statistic function from ggplot2 and indicate the mapping using the appropriate [`after_stat` variable](https://ggplot2.tidyverse.org/reference/aes_eval.html.).

| Chart | Geom | X mapping | Y mapping | Color mapping | Fill mapping | Stat | Position | Coord System | Transform |
|--------|--------|--------|--------|--------|--------|--------|--------|--------|--------|
| A | Bar | nothing | total vaccinations |  | age group | identity |  | polar |  |
| B | point,line | weeks | vaccination rates | age group |  |  |  | cartesian |  |
| C | area | week | vaccination rates |  | age group |  | stack | cartesian |  |
| D | bar | week | vaccination rates | age group |  | identity | dodge | cartesian |  |

: Grammar of graphics mappings for the four charts

3.  Examine each chart and determine whether it effectively communicates changes in vaccine rates over time across different age groups.
    If the chart is inappropriate for that comparison, explain why; if it is appropriate but not optimal, explain what you would change.

    A.   This does not show the time variable at all so that is inapproriate.

    B.  This graph does show the relationship between vaccination rate, time, and age group.
        When looking at this graph you are able to see that changes over time that take place

    C.  This graph does a bad job of showing the changes over time is show a cumulative percentage on the y axis which does not allow you easily distinguish between the contribution between the age groups.

    D.  The bar graph does a great job of the showing the relationship between our variables.
        The relationship however is not as noticeable as the line graph.

4.  Examine each chart and critique the use of Gestalt principles of grouping.
    What principles apply?
    Are they used effectively?

    A.   The coloring of the age groups I believe is misleading as the blues are to similar and I find myself associating them with each other.
        The principle of similarity applies here and it shows how easily things can be associated with each other.

    B.  The coloring of the age groups I believe is misleading as the blues are to similar and I find myself associating them with each other.
        The principle of similarity applies here and it shows how easily things can be associated with each other.

    C.  The coloring of the age groups I believe is misleading as the blues are to similar and I find myself associating them with each other.
        The principle of similarity applies here and it shows how easily things can be associated with each other.

    D.  The coloring of the age groups I believe is misleading as the blues are to similar and I find myself associating them with each other.
        The principle of proximity applies here and it shows how easily things can be associated with each other.
        The principle of continutiy also applies to this graph as we want to connect the bars and and form them into a line.

5.  Consider the "messages" you might want viewers to take away from this data and the visuals you generate.

    a.  Brainstorm 2-3 reasonable messages

    > The message that I would want viewers to take away is that the older generation had the highest vaccination rates.
    > That vaccination rates rose over time.
    > That age is related to vaccination rate.

    b.  What do you think the most important message is from this data set?

    > The most important message of this dataset is how age relates to vaccination rates.

    c.  Which comparisons do you want to make easier for participants to perform?

    > I them to easily be able to compare the vaccination rates of each group with out having the visual miscues of the current coloring

    d.  What obstacles do you expect to encounter designing a chart to facilitate these comparisons?

    > I think that choosing proper coloring will be an issue and formating the lines properly.

6.  Considering your answers to the questions above, create your own version of the chart that minimizes cognitive load and maximizes interpretability.
    Explain the changes that you made and why they improve the chart compared to the ones generated by Claude.

```{r}
#| label: fig-comparing-graphics-figures
#| fig-cap: "COVID Vaccination Rates"
#| fig-subcap: ["Caption 1"]
#| fig-alt: ["This figure is a line graph that shows the relationship between time and vaccination rates by age group. The x - axis shows the week, the y axis shows the vaccination rates, and color indicates the age group."]
#| layout-ncol: 2
#| code-fold: true

library(viridis)
new_chart<- ggplot(vaccination_data, aes(x = week, y = vaccination_rate, 
                                        color = age_group, group = age_group)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_color_viridis(discrete = TRUE, option = "D") +
  labs(title = "Vaccination Rates Over Time by Age Group",
       x = "Week",
       y = "Vaccination Rate (%)",
       color = "Age Group") +
  guides(color = guide_legend(reverse = TRUE)) +
  scale_x_continuous(breaks = 1:12)+
  scale_y_continuous(breaks = seq(0, 100, by = 10),limits = c(0,100))
new_chart
```

# Choose Your Own Adventure

Claude used fake data, but we have access to the [real deal](https://data.cdc.gov/Vaccinations/Weekly-Cumulative-Estimated-Number-of-COVID-19-Vac/ewpg-rz7g/about_data), and I've downloaded a [csv](Weekly_Cumulative_Estimated_Number_of_COVID-19_Vaccinations_Administered_in_Pharmacies_and_Physician_Medical_Offices,_Adults_18_years_and_older,_by_Season_and_Age_Group,_United_States_20251021.csv) to simplify the process.

Of course, the CDC only tracks number of vaccines, not the overall vaccination rate.

1.  Augment this data with census estimates of the population size for each age group.
    In either case, you may need to do a bit of inference/subtraction to get the counts for the age brackets used by the CDC.
    Provide reproducible code that generates a population table for 2023 and 2024 with the appropriate counts for each age bracket provided by the CDC.

    Options:

    -   Use the Census QuickFacts page https://www.census.gov/quickfacts/
    -   Use the [nc-est2024-agesex.xlsx](nc-est2024-agesex.xlsx) population by age and sex file I downloaded from the [US Census](https://www.census.gov/data/tables/time-series/demo/popest/2020s-national-detail.html).

2.  Use the real data to generate the plot you created in 6.
    How does the real data change the appearance/message of the plot?
    Does anything get easier or more difficult?

3.  What annotations would you want to add to your chart?
    You may be able to find relevant information about e.g. approval dates of vaccines at [immunize.org's Vaccine Timeline](https://www.immunize.org/vaccines/vaccine-timeline/).
    Add at least one annotation to your chart, but balance utility with clutter.

4.  How would you modify this chart if your goal was to compare vaccination rates for each age group in 2023 and 2024?
    Explain your answer, highlighting how your modifications would reduce the cognitive load of making the comparison of interest.

5.  How would you use your statistical training to test and visually highlight any statistically significant differences?
    Explain.
    You do not have to actually implement any statistical testing to answer this question.
